---
title: "Design notes and extending synesthR (the rim)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Design notes and extending synesthR (the rim)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## The core / rim architecture

synesthR has a **CRAN-hard core** and an **optional orchestration rim**.

* **Core** — `read -> score -> map -> paint / play / palette / media`. Pure R,
  deterministic given `(text, seed)`, no network, no external software. It must
  pass `R CMD check --as-cran` on a machine with *every* optional package and
  tool absent. The golden snapshot tests fix the numeric output of the mapping;
  that is the package's reproducibility contract.
* **Rim** — optional extensions that orchestrate external tools that *may* exist
  on the user's machine: a local LLM (`ellmer`), Python (`reticulate`), and
  command-line tools (`ffmpeg`, MuseScore). The rim may call the core; **the
  core never calls the rim**. One-directional, absolute.

Every rim function detects its dependency at call time and degrades with an
informative message when it is missing — it never causes a core failure.

## The rim guard pattern

Rim authors reuse four internal helpers rather than hand-rolling detection:

```{r}
.check_tool("ffmpeg")                       # CLI binary on PATH -> path, else abort
.check_pymod("numpy")                       # Python module via reticulate
.check_client("ellmer", probe = my_probe)   # client package + reachable backend
.run_tool("ffmpeg", args = c("-version"))   # system2 wrapper, surfaces exit codes
```

A new extension is then just: guard with the appropriate `.check_*`, do the
work, return a useful result, and document the graceful-absence behaviour.

## Enrichment never alters the core

Optional extensions may *augment* a score (for example
`syn_py_features()` attaches Python-computed valence dynamics to a separate
`enrichment` slot) but never *alter* the deterministic core features or the
mapping. The reproducibility snapshot is identical whether or not any optional
tool is present.

## Suggests discipline

`cld3`, `ellmer`, `gganimate`, `gm`, `magick`, `music`, `reticulate`,
`seewave`, `shiny`, and `soundgen` all live in `Suggests` only and are guarded
with `requireNamespace()`. None ever migrates to `Imports`.
