---
title: "Getting started with synesthR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with synesthR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- badges: start -->
[![R-CMD-check](https://github.com/r-heller/synesthR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/r-heller/synesthR/actions/workflows/R-CMD-check.yaml)
[![pkgdown](https://github.com/r-heller/synesthR/actions/workflows/pkgdown.yaml/badge.svg)](https://r-heller.github.io/synesthR/)
[![CRAN status](https://www.r-pkg.org/badges/version/synesthR)](https://CRAN.R-project.org/package=synesthR)
[![Codecov test coverage](https://codecov.io/gh/r-heller/synesthR/branch/main/graph/badge.svg)](https://app.codecov.io/gh/r-heller/synesthR?branch=main)
[![CRAN downloads](https://cranlogs.r-pkg.org/badges/synesthR)](https://cran.r-project.org/package=synesthR)
[![CRAN downloads total](https://cranlogs.r-pkg.org/badges/grand-total/synesthR)](https://cran.r-project.org/package=synesthR)
[![License: MIT](https://img.shields.io/badge/license-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 6)
```

```{r setup}
library(synesthR)
```

synesthR turns literary text into a single *prosody score* and renders that one
score as both an image and a synchronized piece of audio. The pipeline is
`read -> score -> paint / play / palette`, and it is deterministic: the same
text and seed always produce the same result.

## Read and score

```{r}
text  <- syn_read_text(syn_example_text("en"), lang = "en")
score <- syn_score(text)
score

summary(score)
```

The score holds per-sentence features (valence, length, lexical diversity,
punctuation density) and a smoothed emotional `arc`.

## Paint

```{r paint, fig.alt = "Generative image from the example text."}
syn_paint(score)
```

## Play

```{r}
out <- tempfile(fileext = ".wav")
syn_write_wav(score, out)   # returns the score invisibly, pipe-friendly
file.exists(out)

wave <- syn_play(score)
wave
```

## Palette and theme

The same score yields a colour palette and a matching `ggplot2` theme:

```{r}
pal <- syn_palette(score)
pal
```

Because the image, the audio, and the palette all come from one score, they are
three views of the same analysis.

## Determinism

```{r}
identical(syn_play(score)@left, syn_play(score)@left)
```
