---
title: "From image or audio to palette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From image or audio to palette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(synesthR)
```

`syn_palette()` is a generic: it derives the *same* `syn_palette` object from a
prosody score, a character string, an audio `Wave`, or an image. That makes
palettes interchangeable across sources.

## From text

```{r}
syn_palette("A warm and joyful golden afternoon.", lang = "en")
```

## From audio

Any `tuneR::Wave` works — including the audio synesthR itself generates. The
palette's temperature follows the spectral brightness of the sound (computed
with base R, no extra dependency):

```{r}
w <- tuneR::sine(660, duration = 8000, samp.rate = 8000, bit = 16, pcm = TRUE)
syn_palette(w)
```

## From an image (optional)

The image method uses the optional `magick` package. It downsamples, clusters
colours in Lab space, and orders them by luminance. It runs only when `magick`
is installed:

```{r, eval = requireNamespace("magick", quietly = TRUE)}
img <- magick::image_blank(80, 80, color = "tomato")
syn_palette(img)
```

If `magick` is not installed the method aborts with an informative message
rather than failing the core workflow.

## Turning a palette into a theme

```{r}
pal <- syn_palette(syn_score(syn_read_text("Quiet grey morning by the sea.", lang = "en")))
th <- syn_theme(pal)
class(th)
```
