---
title: "A visual tour of synesthR"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{A visual tour of synesthR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  eval = FALSE,          # the tutorial shows code + pre-rendered figures
  fig.align = "center"
)
# Helper used only by the (evaluated) figure-embedding chunks below.
img <- function(f, w = "100%") knitr::include_graphics(file.path("figures", f))
```

```{r logo, echo = FALSE, eval = TRUE, out.width = "190px"}
img("logo-hex.png")
```

**synesthR** turns a piece of literary text into one shared *prosody score* and
renders that single score as **two synchronized views**: a generative still
image and a piece of self-contained audio. Because the picture and the music
come from the same numbers, they are not two artworks — they are two windows
onto one analysis.

This tour walks through every capability of the package, with a picture for
each. If you just want the shortest path, see
`vignette("getting-started")`; this article is the long, illustrated version.

```{r setup}
library(synesthR)
```

> The figures below were produced by running exactly the code shown. The code
> chunks are not re-evaluated when the vignette builds, so the tutorial renders
> anywhere — even without the optional packages and external tools that some of
> the later sections use.

# The shape of the pipeline

Everything flows one way:

```
text  ──►  syn_read_text()  ──►  syn_score()  ──┬──►  syn_paint()    (image)
                                                 ├──►  syn_play()     (audio)
                                                 └──►  syn_palette()  (colour)
```

A single deterministic mapping sits at the centre: it turns the per-sentence
features into the visual *and* musical parameters. Same text, same seed, same
result — always.

# 1. Reading text

`syn_read_text()` accepts a string, a path to a `.txt` file, or a data frame,
and returns one tidy row per sentence. German, English, and French are
supported. The bundled `syn_example_text()` gives you something to play with.

```{r read}
text <- syn_read_text(syn_example_text("en"), lang = "en")
text
```

```
#> # A tibble: 6 × 3
#>   sentence_id sentence                                            lang
#>         <int> <chr>                                               <chr>
#> 1           1 The morning broke quietly over the harbour.         en
#> 2           2 Grey water lay still against the stones, and the …   en
#> 3           3 A single gull crossed the pale sky, calling once, … en
#> 4           4 She watched from the window and felt the day gath…  en
#> 5           5 Later the wind would rise, and the colours would …  en
#> 6           6 But for now there was only the hush, the patient …  en
```

If the optional **cld3** package is installed you can omit `lang` and let the
language be detected automatically.

# 2. Scoring

`syn_score()` extracts four per-sentence features — valence, length, lexical
diversity, and punctuation density — and a smoothed emotional `arc`.

```{r score}
score <- syn_score(text)
score
summary(score)
```

```
#> <prosody_score> — 6 sentences, lang "en"
#>
#>   n_sentences lang mean_valence valence_range arc_start arc_end mean_length
#> 1           6   en   -0.1666667             3      -0.5     0.5          14
#>   mean_lex_div mean_punct_density
#> 1    0.8900617         0.02743892
```

The per-sentence features are an ordinary tibble you can inspect or plot:

```{r features}
score$features[, c("sentence_id", "valence", "length",
                   "lex_diversity", "punct_density")]
```

```
#>   sentence_id valence length lex_diversity punct_density
#> 1           1      -1      7         0.857        0.0233
#> 2           2       0     14         0.929        0.0267
#> 3           3      -1     12         1.000        0.0429
#> 4           4       0     13         0.923        0.0147
#> 5           5      -1     19         0.737        0.0288
#> 6           6       2     19         0.895        0.0283
```

## The emotional arc

The arc is the smoothed valence trajectory — the backbone that the melody's
contour and the image's gradient both follow.

```{r arc}
plot(score$arc, type = "l")   # or render it nicely, as below
```

```{r arc-fig, echo = FALSE, eval = TRUE, out.width = "95%"}
img("arc.png")
```

# 3. Painting

`syn_paint()` renders the score as a generative still image: an element cloud
whose positions, sizes, hue, and brightness all come from the features, themed
with the score's own palette.

```{r paint}
syn_paint(score)
```

```{r paint-fig, echo = FALSE, eval = TRUE, out.width = "75%"}
img("paint-en.png")
```

### The same machinery, three languages

Feed it different text and the image changes accordingly. Here are the bundled
English, German, and French passages, each painted from its own score:

```{r paint-langs}
syn_paint(syn_score(syn_read_text(syn_example_text("de"), lang = "de")))
syn_paint(syn_score(syn_read_text(syn_example_text("fr"), lang = "fr")))
```

```{r paint-gallery, echo = FALSE, eval = TRUE, out.width = "32%", fig.show = "hold"}
img("paint-en.png"); img("paint-de.png"); img("paint-fr.png")
```

# 4. Playing

`syn_play()` synthesizes the *same* score as audio — a `tuneR::Wave` built from
sine tones with soft harmonics and a click-free envelope. It is fully
self-contained: no external software is ever required.

```{r play}
wave <- syn_play(score)
wave

# Save it to disk (returns the score invisibly, so it pipes):
out <- tempfile(fileext = ".wav")
syn_write_wav(score, out)
```

Here is what that waveform looks like — the audible counterpart of the image
above:

```{r wave-fig, echo = FALSE, eval = TRUE, out.width = "95%"}
img("waveform.png")
```

The valence arc sets the mode (major / minor) and the melodic contour;
punctuation density sets tempo and staccato; sentence length sets note
durations. Picture and sound are two readings of one score.

# 5. Palettes from anything

`syn_palette()` is a generic. It derives the **same** kind of palette object
from a score, a string, an audio `Wave`, or an image — so colours are
interchangeable no matter where they came from.

```{r palette}
syn_palette(score)
```

```
#> <syn_palette> - 5 colours, cool, from prosody_score
#>   #294D00  background
#>   #207900  shadow
#>   #00A65D  mid
#>   #00D5A4  accent
#>   #3BFFEB  highlight
```

```{r pal-score-fig, echo = FALSE, eval = TRUE, out.width = "95%"}
img("pal-score.png")
```

**From text** — score it on the way:

```{r pal-text}
syn_palette("A warm and joyful golden afternoon.", lang = "en")
```

```{r pal-text-fig, echo = FALSE, eval = TRUE, out.width = "95%"}
img("pal-text.png")
```

**From audio** — temperature follows the sound's spectral brightness (base R, no
extra dependency):

```{r pal-audio}
w <- tuneR::sine(660, duration = 8000, samp.rate = 8000, bit = 16, pcm = TRUE)
syn_palette(w)
```

```{r pal-audio-fig, echo = FALSE, eval = TRUE, out.width = "95%"}
img("pal-audio.png")
```

**From an image** — the optional **magick** method clusters colours in Lab space
and orders them by luminance:

```{r pal-image}
img_block <- magick::image_blank(80, 80, color = "tomato")
syn_palette(img_block)
```

```{r pal-image-fig, echo = FALSE, eval = TRUE, out.width = "95%"}
img("pal-image.png")
```

## Wearing the palette: `syn_theme()`

Any palette becomes a `ggplot2` theme — the darkest role becomes the
background, the lightest the text and lines. So *any* plot can be dressed in a
text's colours:

```{r theme}
library(ggplot2)
pal <- syn_palette(score)
ggplot(mtcars, aes(wt, mpg, colour = mpg)) +
  geom_point(size = 3) +
  scale_colour_gradientn(colours = pal$colors) +
  syn_theme(pal)
```

```{r theme-fig, echo = FALSE, eval = TRUE, out.width = "95%"}
img("theme-demo.png")
```

# 6. Animating the arc (optional)

With the optional **gganimate** package, `syn_animate()` reveals the element
cloud sentence by sentence, so the emotional arc unfolds in time:

```{r animate}
anim <- syn_animate(score)
gganimate::animate(anim, nframes = 48, fps = 12)
```

```{r animate-fig, echo = FALSE, eval = TRUE, out.width = "55%"}
img("animate.gif")
```

# 7. The media interchange container

Functions in the suite pass objects as `syn_media` — a small wrapper carrying
the content, a type tag, and metadata. It is the common currency that lets
text, images, audio, and scores flow between functions.

```{r media}
m <- as_syn_media(score)
m
media_type(m)
```

```
#> <syn_media> of type "score"
#>   meta: "lang"
#> media_type: "score"
```

`as_syn_media()` has methods for character (text), `magick-image` (image),
`Wave` (audio), and `prosody_score` (score).

# 8. The optional "rim": assistive extensions

synesthR has a CRAN-hard **core** (everything above) and an optional
orchestration **rim** that talks to tools which *may* exist on your machine. The
rim never alters the deterministic core; every entry point degrades with a
helpful message when its dependency is absent. (See
`vignette("extending")` for the design rules.)

## A local-LLM reading

With the optional **ellmer** package and a reachable local model (e.g. Ollama),
`syn_interpret()` returns a short prose "reading" of *why* the image and music
came out as they did. It is sent only the features — never your raw text — and
is advisory, never part of the mapping.

```{r interpret}
syn_interpret(score)                    # a paragraph of interpretation
syn_suggest(score, target = "tempo")    # an advisory tweak; never auto-applied
syn_llm_condense(text)                  # opt-in: shorten very long input first
```

## Python-computed enrichment

With **reticulate** and **numpy**, `syn_py_features()` attaches auxiliary
valence dynamics (gradient and cumulative trajectory) to a separate
`enrichment` slot — additive only; the core features and the reproducibility
snapshot are untouched.

```{r py}
score2 <- syn_py_features(score)
score2$enrichment$py_features
```

## A richer engraved score

With the optional **gm** package and a MuseScore install, `syn_play_gm()` builds
an engraved, notated version of the same notes for richer playback:

```{r gm}
music <- syn_play_gm(score)
gm::show(music)        # engrave / play (requires MuseScore)
```

## Video export

`syn_export_video()` is a documented stub in this release: it demonstrates the
CLI-orchestration pattern (it probes for `ffmpeg`) but intentionally aborts with
guidance — use `syn_paint()` + `syn_write_wav()` for now.

# 9. The interactive app

`syn_run_app()` launches a small Shiny app: paste text, pick a language, and see
the image and palette while hearing the audio — the three core views side by
side. The LLM panel lights up only if an `ellmer` backend is reachable.

```{r app}
syn_run_app()
```

It is, in essence, this whole tour made interactive: every panel is one of the
functions you have just seen, driven by the one shared score.

# 10. Determinism, the whole point

Because the core is a pure, seeded mapping, the same text always yields the
same image, the same audio, and the same palette:

```{r determinism}
identical(syn_play(score)@left, syn_play(score)@left)
#> [1] TRUE
```

That reproducibility is the package's contract: image, sound, and colour are
three faithful views of one analysis, today and next year.

# Where to go next

* `vignette("getting-started")` — the five-minute version.
* `vignette("palette-from-media")` — palettes from images and audio in depth.
* `vignette("extending")` — the core/rim architecture and how to add your own
  rim extension.
```
