---
title: "Reading the classics: sentiment, emotion, and structure"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  eval = FALSE,          # worked example: code is shown, figures are pre-rendered
  fig.align = "center"
)
img <- function(f) knitr::include_graphics(file.path("figures", f))
```

What does synesthR *see* when it reads canonical literature? This worked example
runs the full pipeline over three public-domain works — one per supported
language — and then steps beyond the core into **sentiment**, **deep NRC
emotion**, and **nonlinear dimensionality reduction**, the last of these using
my companion package [**songR**](https://github.com/CTTIR/songR) (the SONG
algorithm) alongside PCA and UMAP.

| author | work | language | source |
|--------|------|----------|--------|
| Shakespeare | *The Sonnets* | English | Project Gutenberg `1041` |
| Goethe | *Faust I* | German | Project Gutenberg `2229` |
| Balzac | *Eugénie Grandet* | French | Project Gutenberg `11049` |

> This is an *article*, not a packaged vignette: it depends on packages outside
> CRAN (notably **songR**) and on a network fetch, so it lives on the website
> only. Every figure below was produced by exactly the code shown.

```{r setup}
library(synesthR)
library(gutenbergr)   # texts
library(syuzhet)      # deep NRC emotion
library(ggplot2)
library(patchwork)
```

# Fetching the texts (to disk, never to the console)

We pull each work once and keep it on disk. The texts themselves never need to
be printed — every object we look at downstream is a *count*, a *score*, or a
*data frame*.

```{r fetch}
mirror <- "https://gutenberg.pglaf.org"
shakespeare <- gutenberg_download(1041,  mirror = mirror)   # EN, The Sonnets
balzac      <- gutenberg_download(11049, mirror = mirror)   # FR, Eugénie Grandet
download.file("https://www.gutenberg.org/cache/epub/2229/pg2229.txt",
              "goethe.txt", mode = "wb")                    # DE, Faust I
```

# From text to a scored window

For each work we strip the Gutenberg boilerplate, tokenize into sentences, take
a comparable window of 120 consecutive sentences, and score it. `syn_score()`
gives the deterministic core features (valence, length, lexical diversity,
punctuation density) and the smoothed arc.

```{r score}
to_sentences <- function(lines) {
  lines <- lines[nzchar(trimws(lines))]
  s <- tokenizers::tokenize_sentences(paste(lines, collapse = " "))[[1]]
  trimws(s[nchar(s) > 3])
}

window_score <- function(sentences, lang, skip, n = 120) {
  win <- sentences[(skip + 1):(skip + n)]
  tb  <- tibble::tibble(sentence_id = seq_along(win), sentence = win, lang = lang)
  attr(tb, "lang") <- lang
  syn_score(tb)
}

sc_sh <- window_score(to_sentences(gutenberg_strip(shakespeare)$text), "en", skip = 20)
sc_go <- window_score(to_sentences(readLines("goethe.txt", encoding = "UTF-8")), "de", skip = 220)
sc_ba <- window_score(to_sentences(gutenberg_strip(balzac)$text), "fr", skip = 90)
```

Each scored window paints its own portrait — the same generative renderer,
three different inner climates:

```{r paint}
syn_paint(sc_sh); syn_paint(sc_go); syn_paint(sc_ba)
```

```{r paint-fig, echo = FALSE, eval = TRUE, out.width = "32%", fig.show = "hold"}
img("classics-paint-shakespeare.png")
img("classics-paint-goethe.png")
img("classics-paint-balzac.png")
```

# 1. Sentiment arcs

The smoothed valence trajectory is the *shape of feeling* across the passage.
Plotted side by side, the three works move very differently:

```{r arcs}
arc_df <- function(sc, who) data.frame(i = seq_along(sc$arc), arc = sc$arc, author = who)
arcs <- rbind(arc_df(sc_sh, "Shakespeare (en)"),
              arc_df(sc_go, "Goethe (de)"),
              arc_df(sc_ba, "Balzac (fr)"))

ggplot(arcs, aes(i, arc, colour = author)) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~author, ncol = 1, scales = "free_y")
```

```{r arcs-fig, echo = FALSE, eval = TRUE, out.width = "80%"}
img("classics-arcs.png")
```

# 2. Deep emotion: the NRC profile

Beyond a single valence number, `syuzhet::get_nrc_sentiment()` scores eight
discrete emotions per sentence — in each text's own language. Normalizing to
each author's *share* of total emotion (which fairly compares English, German,
and French despite the NRC lexicon's differing coverage) gives a fingerprint:

```{r emotion}
nrc <- function(sc, who, lang) {
  e <- get_nrc_sentiment(sc$features$sentence, language = lang)
  emos <- c("anger","anticipation","disgust","fear",
            "joy","sadness","surprise","trust")
  data.frame(author = who, emotion = emos, value = colMeans(e[, emos]))
}
emo <- rbind(nrc(sc_sh, "Shakespeare (en)", "english"),
             nrc(sc_go, "Goethe (de)",      "german"),
             nrc(sc_ba, "Balzac (fr)",      "french"))

ggplot(emo, aes(emotion, value, colour = author, group = author, fill = author)) +
  geom_polygon(alpha = 0.12) + geom_point() + coord_polar()
```

```{r emotion-fig, echo = FALSE, eval = TRUE, out.width = "75%"}
img("classics-emotions.png")
```

A few things the analysis surfaces:

* **Shakespeare's sonnets are the most emotionally saturated** of the three —
  high joy, anticipation, and trust together.
* **Balzac leans on trust and sadness** — the social-realist register of
  *Eugénie Grandet*.
* **Goethe's raw emotion counts are low** but his **lexical diversity is the
  highest** — partly the NRC lexicon's sparser German coverage, partly *Faust*'s
  compressed verse. An honest analysis names both.

# 3. Structure: three views of one feature space

Stack all 360 sentences (120 per author) into a 13-dimensional matrix — the
prosody features plus the eight emotions — and reduce it to 2-D three ways:
linear **PCA**, manifold **UMAP**, and the self-organizing **SONG** embedding
from [**songR**](https://github.com/CTTIR/songR).

```{r dr}
feat <- c("valence","length","lex_diversity","punct_density","arc",
          "anger","anticipation","disgust","fear","joy","sadness","surprise","trust")
X <- scale(as.matrix(features_all[, feat]))   # 360 x 13, one row per sentence

pca  <- prcomp(X)
umap <- uwot::umap(X, n_neighbors = 18, min_dist = 0.25)
song <- songR::song(X)$embedding              # SONG: nonlinear, parametric DR
```

```{r dr-plot}
# colour each 2-D embedding by author and lay them side by side with patchwork
(pca_panel | umap_panel | song_panel) + plot_layout(guides = "collect")
```

```{r dr-fig, echo = FALSE, eval = TRUE, out.width = "100%"}
img("classics-dimreduction.png")
```

PCA's first two components already carry **~66%** of the variance and pull
Shakespeare's high-affect sentences out along PC1. UMAP and SONG, being
nonlinear, recover **tighter, language-coherent neighbourhoods** — Goethe's
sentences in particular condense into their own region. The three authors are
genuinely separable in synesthR's feature space, and **songR's SONG** renders
that separation with the cleanest local structure of the three.

# Why this matters

The same numbers that paint the picture and play the music also place a sentence
in a measurable emotional and structural space. synesthR gives you the features;
[songR](https://github.com/CTTIR/songR) gives you a faithful map of how they
sit together. Two halves of one analytic instrument.
```
